Determinism, Complexity, and Predictability in Computer Performance

### 2014-07-24 wpe.tex
airplane edits

Previous approaches to measuring temporal complexity via the Shannon
entropy rate \cite{Shannon1951, mantegna1994linguistic} required
categorical data: $x_i \in \mathcal{S}$ for some finite or countably
infinite \emph{alphabet} $\mathcal{S}$.  Data taken from real-world
systems are, however, effectively\footnote{Measurements from
  finite-precision sensors are discrete, but data from modern
  high-resolution sensors are, for the purposes of entropy
  calculations, effectively continuous.}  real-valued.  To analyze
real-valued data using a method that requires categorical values, one
must discretize the data---typically by binning.  Unfortunately, this
is rarely a good solution, as the binning of the values introduces
spurious dynamics~\cite{bollt2001}.  The field of symbolic dynamics
offers discretization methods that do not disturb the intrinsic
behavior.  These methods are, however, fragile in the face of noise;
worse yet, they require knowledge of the underlying system.  This is
not useful in our context, where we want to measure the entropy to
quantify predictability of an arbitrary time-series.

@@

In practice, calculating permutation entropy and weighted permutation
entropy involves choosing a good value for the word length $\ell$. The
primary consideration in that choice is that the value be large enough
that forbidden ordinals are discovered, yet small enough that
reasonable statistics over the ordinals are gathered.  If an average
of 100 counts per ordinal is considered to be sufficient, for
instance, then $\ell = \operatornamewithlimits{argmax}_{\hat{\ell}} \{
N \gtrapprox 100 \hat{\ell}! \}$.  In the literature, $3 \le \ell \le
6$ is a standard choice---generally without any formal justification.
In theory, the permutation entropy should reach an asymptote with
increasing $\ell$, but that can require an arbitrarily long time
series. In practice, what one should do is calculate the
\emph{persistent} permutation entropy by increasing $\ell$ until the
result converges, but data length issues can intrude before that
convergence is reached.  We used this approach to choose $\ell = 6$
for the experiments in the following section.  This value represents a
good balance between accurate ordinal statistics and finite-data
effects.

>>>>

Previous approaches to measuring temporal complexity via the Shannon
entropy rate \cite{Shannon1951, mantegna1994linguistic} required
categorical data: $x_i \in \mathcal{S}$ for some finite or countably
infinite \emph{alphabet} $\mathcal{S}$.  Data taken from real-world
systems are, however, effectively\footnote{Measurements from
  finite-precision sensors are discrete, but data from modern
  high-resolution sensors are, for the purposes of entropy
  calculations, effectively continuous.}  real-valued.  To analyze
real-valued data using a method that requires categorical values, one
must discretize the data---typically by binning.  Unfortunately, this
is rarely a good solution, as the binning of the values introduces
spurious dynamics~\cite{bollt2001}.  The field of symbolic dynamics
offers discretization methods that do not disturb the intrinsic
behavior.  These methods are, however, fragile in the face of noise;
worse yet, they require knowledge of the underlying system.  This is
not useful in our context, where the target of study is experimental
time-series data.

@@

In practice, calculating permutation entropy and weighted permutation
entropy involves choosing a good value for the word length $\ell$. The
primary consideration in that choice is that the value be large enough
that forbidden ordinals are discovered, yet small enough that
reasonable statistics over the ordinals are gathered.  If an average
of 100 counts per ordinal is considered to be sufficient, for
instance, then $\ell = \operatornamewithlimits{argmax}_{\hat{\ell}} \{
N \gtrapprox 100 \hat{\ell}! \}$.  In the literature, $3 \le \ell \le
6$ is a standard choice---generally without any formal justification.
In theory, the permutation entropy should reach an asymptote with
increasing $\ell$, but that can require an arbitrarily long time
series. In practice, what one should do is calculate the
\emph{persistent} permutation entropy by increasing $\ell$ until the
result converges, but data length issues can intrude before that
convergence is reached.  We used this approach, which represents a
good balance between accurate ordinal statistics and finite-data
effects, to choose $\ell$ values for the experiments in the following
section.
